On Extensibility and Personalizability of Multi-Modal Trip Planning

### 2016-03-03 intro.tex
Broke paper.tex in pieces.

\def\year{2016}\relax
%File: formatting-instruction.tex
\documentclass[letterpaper]{article}
\usepackage{aaai16}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}

\usepackage[pdftex]{hyperref}
\usepackage{amsmath}
\usepackage{color}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{ccicons}
\usepackage{todonotes}
\usepackage{amsfonts}
\usepackage{framed}

\frenchspacing
\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}
\pdfinfo{
/Title (On Extensibility and Personalizability of Multi-Modal Trip Planning)
/Author (Put All Your Authors Here, Separated by Commas)}
\setcounter{secnumdepth}{0}

\newcommand{\aux}{\mathit{aux}}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\cA}{\mathcal{A}}
\newcommand{\cU}{\mathcal{U}}
\newcommand{\cP}{\mathcal{P}}
\newcommand{\tit}[1]{\textit{#1}}
\newcommand{\IFF}{\textit{iff}}
\newcommand{\IF}{\textit{if}}
\newcommand{\PCF}{\mathit{PCF}}
\newcommand{\walk}{\mathit{walk}}
\newcommand{\bike}{\mathit{bike}}
\newcommand{\car}{\mathit{car}}
\newcommand{\public}{\mathit{public}}
\newcommand{\taxi}{\mathit{taxi}}
\newcommand{\gas}{\mathit{gas}}
\newcommand{\wait}{\mathit{wait}}
\newcommand{\SUM}{\mathit{sum}}
\newcommand{\MAX}{\mathit{max}}
\newcommand{\MIN}{\mathit{min}}
\newcommand{\AVG}{\mathit{avg}}

\newcommand{\secref}[1]{Section~\ref{sec:#1}}
\newcommand{\figref}[1]{Figure~\ref{fig:#1}}
\newcommand{\tblref}[1]{Table~\ref{tbl:#1}}
\newcommand{\algref}[1]{Algorithm~\ref{alg:#1}}
\newcommand{\thmref}[1]{Theorem~\ref{thm:#1}}
\newcommand{\propref}[1]{Proposition~\ref{prop:#1}}
\newcommand{\corref}[1]{Corollary~\ref{cor:#1}}
\newcommand{\eqtref}[1]{Equation~\ref{eqt:#1}}

 \begin{document}
% The file aaai.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\title{On Extensibility and Personalizability of Multi-Modal Trip Planning}
%\author{
%	Xudong Liu\\
%	Department of Computer Science\\
%	University of Kentucky\\
%	329 Rose Street\\
%	Lexington, KY 40506\\
%	USA
%	\And
%	PARC People\\
%	Palo Alto Research Center\\
%	3333 Coyote Hill Road\\
%	Palo Alto, CA 94304\\
%	USA
%}
\maketitle
\begin{abstract}
	Trip planning is a practical task that has drawn
	extensive attention from the AI planning and scheduling community
	and the industrial/commercial sectors.
	In this paper, we consider the setting of the \tit{multi-modal}
	trip planning, where users can exploit different transportation
	modes, such as walking, biking, public transit, and taxi.
	In such a context, it would be of benefit if the
	user was able to extend the \tit{cost base}, including traveling time 
	and fare, of the planner, and to personalize the planner according
	to her own constraints and preferences.
	To this end, we designed and developed a hipergraph-based multi-modal
	trip planner that allows users to upload auxiliary cost metrics (e.g.,
	crime rates), to specify constraints as a theory in the 
	\tit{linear temporal logic}, and to express preferences as a 
	\tit{preferential cost function}.
\end{abstract}

\section{Introduction}
Trip planning, as an application of planning and scheduling,
has seen substantive implementations by researchers and
developers\cite{bast2015route}.
Some of the planning systems are \tit{multi-modal}; that is,
combining distinct transportation modes, the trip planners
compute optimal routes from sources to destinations.
This notion of ``optimality" generally refers to the computed routes
having minimal total time or total fare.

@@

\section{Extensibility}
%\begin{itemize}
%	\setlength\itemsep{1pt}
%	\item Explain in detail what extensibility means.
%	\item Describe the input.
%	\item Describe how we incorporate the auxiliary data into the planner.
%\end{itemize}
Allowing users to upload their own data sets of interests is an important
step towards customization of a trip planner.
We designed a framework where a user can upload auxiliary cost 
metric data (e.g., crime
statistics and pollution data) into the planner,
and the planner will compute an optimal route accordingly.

The user-created data are the auxiliary data that is represented as pairs
of latitude and longitude degrees.
To merge these lat-long pairs into the planner, we performed a neighborhood
search to calculate the total score of auxiliary data for each 
lat-long pair already in our planner.
It might be of strong interest to some user for our planning
system to take care of criminal statistics so that some level of safety
of the resulting routes is guaranteed.
For instance, a user traveling through the downtown area of San Francisco
around midnight may want to upload a data set of crimes (cf. \figref{sf_crime}), 
and express her constraints and preferences in hope of a safer trip plan.
Note that in \figref{sf_crime} bad areas are the colored circles, whose
integer labels represent numbers of crimes in corresponding areas.

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.4\textwidth]{figs/sub_sf_crime_July2015.png}
  \caption{Crime rates in San Francisco\label{fig:sf_crime}}
\end{figure}

Formally, let $\cU$ be the uploaded auxiliary dataset of lat-long pairs, 
$N=(x_N,y_N)$ a node in our planner and $r$ the effective radius.
We first compute the set $\cP \subseteq A$ of auxiliary points such that,
for every node $P \in \cP$, the Euclidean distance between $N$ and $P$
is within $r$.
Then, the score of auxiliary data for $N$ with respect to $\cP$ and $r$,
denoted by $S(N,\cP,r)$, is computed as follows.

\begin{equation*}
	S(N,\cP,r) = \sum_{\substack{P \in \cP}} 
		\Big(1 - \frac{\sqrt{(x_N-x_P)^2+(y_N-y_P)^2}}{r}\Big)
\end{equation*}

Now we turn to \figref{aux} for an instance to show how auxiliary data
are integrated into the graph using the equation above.
In \figref{aux}, we have the green nodes denoting the graph nodes, 
and the red nodes the new auxiliary nodes that represent locations
of criminal events.
For node $N$, there are three red nodes within its neighborhood of
radius of $r$; thus, we have $\cP=\{P_1,P_2,P_3\}$.
We assume $r$ is 100 feet, and the distances from $N$ to $P_1$,
$P_2$ and $P_3$ are 25, 80 and 90 feet, respectively.
Then, the crime score of $S(N,\cP,r)$ is $1.05$.

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.4\textwidth]{figs/aux.pdf}
  \caption{Integrating uploaded data\label{fig:aux}}
\end{figure}

\section{Personalizability}
Personalizability consists of two aspects: constraints and preferences.
From the viewpoint of the planner,
constraints, also referred to as hard constraints, are statements that the planner
has to satisfy during the planning process; whereas preferences, also called
soft constraints, are specifications that the planner will need to optimize.
We formulated constraints using linear temporal logic (LTL) and preferences as
a preferential cost function (PCF), and implemented our planner leveraging the
widely-used graph search algorithm the A*.

\subsection{Constraints}
%\begin{itemize}
%	\setlength\itemsep{1pt}
%	\item Describe syntax and semantics of LTL in the setting of trip planning.
%	\begin{itemize}
%		\setlength\itemsep{0pt}
%		\item Specifically, we need to point out the ``after" temporal connective,
%					and we need to describe the semantics: what is a plan, why we can
%					eliminate actions and put them as state attributes instead, etc.
%	\end{itemize}
%\end{itemize}
As constraints in the setting of trip planning are often declarative and
temporal, our choice of LTL is straightforward.
We now give a brief review of linear temporal logic (LTL).
Let $f$ be a propositional formula over a finite set $L$ of Boolean variables.  
LTL formulas are defined recursively as follows.
\begin{equation}
	\varphi = f | \varphi_1 \land \varphi_2 | \varphi_1 \lor \varphi_2 | \neg \varphi | 
		\bigcirc \varphi |	\Box \varphi | \Diamond \varphi | \varphi_1 \cA \varphi_2
\end{equation}
Note that we have $\varphi_1 \cA \varphi_2$, and it means that
``$\varphi_2$ holds right after $\varphi_1$ holds."

A natural constraint in trip planning could be ``In this trip I will not drive a car 
after biking or taking the public transit."
In LTL, such constraint can be translated into an LTL formula
\begin{equation}
\label{eqt:ex}
	\psi = ((M=b) \lor (M=p)) \,\cA\, (\Box (\neg (M=c))).
\end{equation}

As the actions in trip planning is limited to taking different transportation modes,
in our definition of the semantics of LTL
these actions are subsumed into the interpretations of $L$, or \tit{states}.
The semantics of LTL is defined with regard to trajectories of states. 
Let $\sigma$ be a trajectory of states $S_0,a_1,S_1,\ldots,a_n,S_n$, and
$\sigma[i]$ a suffix $S_i, a_{i+1}, S_{i+1}, \ldots,a_n,S_n$.  We have
\begin{align*}
	\sigma \models f \;\; &\IFF \;\; S_0 \models f,\\
	\sigma \models \varphi_1 \land \varphi_2 \;\; &\IFF \;\; \sigma \models \varphi_1 \; and \; \sigma \models \varphi_2,\\
	\sigma \models \varphi_1 \lor \varphi_2 \;\; &\IFF \;\; \sigma \models \varphi_1 \; or \; \sigma \models \varphi_2,\\
	\sigma \models \neg \varphi \;\; &\IFF \;\; \sigma \not \models \varphi,\\
	\sigma \models \bigcirc \varphi \;\; &\IFF \;\; \sigma[1] \models \varphi,\\
	\sigma \models \Box \varphi \;\; &\IFF \;\; \forall 0 \leq i \leq n (\sigma[i] \models \varphi),\\
	\sigma \models \Diamond \varphi \;\; &\IFF \;\; \exists 0 \leq i \leq n (\sigma[i] \models \varphi),\\
	\sigma \models \varphi_1 \cA \varphi_2 \;\; &\IFF \;\; \forall 0 \leq i < n (\IF \; \sigma[i] \models \varphi_1, \sigma[i+1] \models \varphi_2).
\end{align*}

For example, we are given an LTL constraint $\psi$ and three trajectories ($\sigma_1$, $\sigma_2$ and $\sigma_3$)
as shown in \figref{trjs}.
Clearly, we have $\sigma_1 \models \psi$ because, after public transit in $S_2$ and $S_3$, 
traveling by car has never been taken place.
Moreover, we have $\sigma_2 \not \models \psi$ because we have $M=c$ hold in $S_6$ and $S_7$
after having $M=p$ hold in $S_2$ and $S_5$.
Finally, we know $\sigma_3 \models \psi$, as the mode is always neither biking nor public transit.

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.4\textwidth]{figs/trajectories.pdf}
  \caption{State transition diagram\label{fig:trjs}}
\end{figure}

\subsection{Preferences}
%\begin{itemize}
%	\setlength\itemsep{1pt}
%	\item Describe our preferential cost function (PCF).
%	\begin{itemize}
%		\setlength\itemsep{0pt}
%		\item Describe why and how we break down time and money.
%		\item Describe why and how we design coefficients to these time and money pieces.
%		\item Describe why and how we design the two ratios: dollars per hour and dollars per aux.
%	\end{itemize}
%\end{itemize}

A state is described as a set of \tit{state variables}.
The state variables of a state $S$ include the transportation mode $M$ that led to $S$,
time $T$ spent so far per mode (e.g., $T_\bike$ for biking and $T_\public$ for
public transit), fare $D$ spent so far per mode (e.g., $D_\gas$ for driving and
$D_\taxi$ for taking a cab), and variables related to the auxiliary data once uploaded.
These extra data related variables are metrics such as the sum ($A_\SUM$),
the maximum ($A_\MAX)$, the minimum ($A_\MIN$), and the average ($A_\AVG$) data along the path.

We focused on weighted functions over state variables and
designed the cost function, called preferential cost function (PCF), that guides the
graph-based search engine in our trip planner as follows.
\begin{equation}
	\begin{aligned}
		\PCF(S) =& \beta_1 * (\alpha_\walk \cdot T_\walk + \alpha_\bike \cdot T_\bike + \ldots) \\
								&+ (D_\gas + D_\public + \ldots) \\
								&+ \beta_2 * (A_\SUM + \ldots),
	\end{aligned}
	\label{eqt:pcf}
\end{equation}
where $\alpha_i$ are real numbers representing the relations among different time pieces,
and $\beta_1$ ($\beta_2$) is the ratio that essentially describes how much in dollars a user would pay to
save an hour (an auxiliary data, respectively).

\subsubsection{Preference Elicitation}
To gather these coefficients ($\alpha_i$'s and $\beta_i$'s) in our $\PCF$, we designed interface to
elicit these numbers from the user.
It asks the user questions and collect answers from the user to derive the coefficients.
These questions are as follows.
\begin{enumerate}
	\item One hr of walking $=$ \underline{\hspace{1cm}} hrs of driving?
	\item One hr of biking $=$ \underline{\hspace{1cm}} hrs of driving?
	\item One hr of public transit $=$ \underline{\hspace{1cm}} hrs of driving?
	\item One hr of taxi $=$ \underline{\hspace{1cm}} hrs of driving?
	\item How much in dollars would you pay to save an hour in traveling?
	\item How much in dollars would you pay to avoid an auxiliary event (e.g., crime) in traveling?
\end{enumerate}
For instance, Alice, an agent, answers 3, 2, 0.25 and 0.5 to the first four questions in the list above.
Intuitively, the numbers indicate that she prefers public transit the most, followed by
taxi, driving, biking and walking, in order.
We show how we can now derive $\alpha_i$'s in \eqtref{pcf}.
We start with setting $\alpha_\car=1$.
Now, since one hour of walking is equivalent to 3 hours of walking,
we have $\alpha_\walk \times 1 = \alpha_\car \times 3$;
hence, we derive $\alpha_\walk = 3$.
Similarly, we have $\alpha_\bike = 2$, $\alpha_\public = 0.25$,
and $\alpha_\taxi = 0.5$.
Computing $\beta_1$ and $\beta_1$ is straightforward.
Indeed, function $\PCF$ with the input of time, fare and
auxiliary metric pieces boils down to monetary cost,
and the planner computes the best path by optimize
based on this overall monetary cost in the searching
process.

\subsection{Reasoning with Constraints and Preferences}
%\begin{itemize}
%	\setlength\itemsep{1pt}
%	\item Describe why and how we integrate constraints and preferences into A*.
%\end{itemize}

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.4\textwidth]{figs/Astar.pdf}
  \caption{Adjusted A*\label{fig:astar}}
\end{figure}

We leveraged the widely-used A* search algorithm on top of our high-performance graph search
engine (cf. \figref{astar}).  The A* algorithm incorporates the following cost function.
\begin{equation}
	f(S) = g(S) + h(S),
\end{equation}
where $g(S)$ is the cost of an optimal trip from the initial state to $S$, and
$h(S)$ is an admissible estimate of the cost of an optimal trip from $S$ to goal.

To prune the search space, we check satisfiability of the temporal constraints in LTL
at expansion of the search tree.
To guide the search engine, we set $g(S)=\PCF(S)$ and $h(S)$ the minimum estimate among
all available modes in $S$.

%\section{User Study}
%\begin{itemize}
%	\setlength\itemsep{1pt}
%	\item Describe the set up of our experiments on AMT and explain why: because we want to evaluate the
%				feasibility of the three features.
%	\item Describe the results.
%\end{itemize}
\section{Implementation and Results}
%\begin{itemize}
%	\setlength\itemsep{1pt}
%	\item Describe the system structure.
%	\item Describe the input to the system.
%	\item Describe the output from the system.
%\end{itemize}

\subsection{System Architecture}
\begin{figure*}[!ht]
  \centering
    \includegraphics[width=0.8\textwidth]{figs/system.pdf}
  \caption{System Overview\label{fig:system}}
\end{figure*}
We designed and implemented a multi-modal trip planning
system (cf. \figref{system}) based on a 
high-performance graph search engine.
The planner allows user uploads, as well as declarative
constraints and preferences.
We now describe the structure of the planning system.

The trip planner takes two types of data as input:
static data and user-specified request.
The static input includes Map Data and Transit
Data.
Map Data describes the map, a directed graph where 
nodes are street corners, bus stops and train stations.
Transit Data is a set of schedules for the buses and trains
On the other hand, a user provides her request, composed of 
three parts.
First, the user enters \tit{from} and \tit{to} locations 
on the map together with day and time of the start of the trip.  
Second, the user may upload her auxiliary metric dataset, e.g., crime rates.  
Lastly, the user specifies her constraints in LTL and preferences as a $\PCF$.
For example, the constraint 
could be ``never walk through a bad neighborhood."  Given these 
inputs, our planner computes an optimal path satisfying 
all the constraints and optimizing the preferences.
Note that the request from the user is encapsulated into
a JSON object.

For instance, the JSON object for the constraint $\psi$ in \eqtref{ex}
is shown in \figref{exJSON}.
\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/exampleJSON.png}
  \caption{JSON object for constraint $\psi$ in \eqtref{ex}\label{fig:exJSON}}
\end{figure}

\subsection{Results}
We present and analyze resulting routes for three agents:
Alice, Bob, and Cal.
This is assuming no auxiliary metric datasets uploaded
so that the agents focus on time and fare.
Fixing their where and when information, we show how
agents' different constraints and preferences affect
their optimal routes computed by our planner.
Their where and when are set so that they all plan to
travel from San Jose International Airport (SJC) in
San Jose, to Pier 39 in San Francisco.
Note that the natural constraint $\psi$ in \eqtref{ex} 
is implicitly imposed on all cases.

Agent Alice has only one constraint that she does not 
have a car; thus, driving to her is never available during this trip.
Per her preferences, Alice provides her thoughts as earlier;
that is, public transportation is preferred to taxi, which is
better than biking, preferred to walking.
What's more, she decides that she would sacrifice thirty dollars
to save one hour during the trip.

The resulting route for Alice is included in \figref{alice}.
The solution takes 2 hours 7 minutes and 18 dollars 94 cents.
It is optimal in the sense the combined value of time
and fare is minimal among all paths from SJC to Pier 39.

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/result_Alice.pdf}
  \caption{Resulting route for Alice\label{fig:alice}}
\end{figure}

Like Alice, Bob is constrained that he will not drive a car in his travel,
and his dollar per hour is thirty.
Moreover, Bob makes his mind to have some workout with his bike and
dictates that he will bike for between one and two hours.
He expresses his preferences: biking and public transit are the most
preferred, next is taxi, and the least preferred is walking.
Similarly, he has done so by answering the aforementioned elicitation questions.

The result for Bob is depicted in \figref{bob}.
It spans 2 hours 57 minutes in time with the fare of 29 dollars 17 cents.
It is so, seemingly worse than what Alice achieved, only because Bob has
the constraint that he will for sure bike for one to two hours, and
the preferences that put biking the most satisfying mode.

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/result_Bob.pdf}
  \caption{Resulting route for Bob\label{fig:bob}}
\end{figure}

Agent Cal will also not use a car in the trip.
He affirms that time is greatly valuable and one hour is measured
to 500 dollars, although his budget is restricted to 50.
Cal's preferences are that the most preferred are public transit
and taxi, and the next preferred are walking and biking that
are equivalent.

Refer to \figref{cal} for the optimal path for Cal.
This route, stretching 1 hour 48 minutes, is the most time-saving one 
compared with the previous two, at the price of 49 dollars 91 cents.
This is due to the constraint that Cal can spend up to 50 dollars,
as well as his preferences and vale of time being high.

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/result_Cal.pdf}
  \caption{Resulting route for Cal\label{fig:cal}}
\end{figure}

\subsubsection{Auxiliary Metric}
When the user of the planner is interested in metrics other than
the ones offered already (i.e., time and fare), she might discover
new metrics (e.g., crime rates and pollution statistics), upload
them into the planner, and retrieve optimal plans taking these
metrics into account.  One scenario of this approach is the following.

Our agent needs to travel without a car across San Francisco downtown at night.
For her, safety is important.
Having found the crime statistics for the area, the agent uploads
the data as a new auxiliary metric into the map.
By specifying that she will never walk through a neighborhood
with more than fifteen crimes over the last month, and that
she would sacrifice a quarter to avoid one crime incident,
the agent tells the planner to come up with a relatively safe
route. An example is shown in \figref{crime}, where the agent
needs to start at the east of downtown and travel across the area
to arrive at the west side.  
The computed path is represented by the line colored by black,
blue and green, denoting taxi, public transit and biking, respectively.
Clearly, this path routes away from crime-heavy areas and achieves
optimality in that the combined metrics -- time, fare and crime rates,
uploaded and personalized by the user -- is minimal among all possibilities.

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/crime.pdf}
  \caption{Optimal route considering crime rates\label{fig:crime}}
\end{figure}

\section{Conclusion and Future Work}
In multi-modal trip planning, it is vital for the planner to enable
users to integrate new metrics of interest into the existing map,
and to express constraints and preferences over which optimal
routes are computed.
In our work, we designed and built a graph-search based, extensible
and personalizable multi-modal trip planner that utilizes linear
temporal logic and preferential cost
function to model user constraints and preferences.

As a note on the future work, we plan to explore techniques in planning
for computing multiple routes that are diverse with bounded difference of costs.
Continuing this direction, we intend to study the problem of learning the
$\PCF$ coefficients using the observations of the decisions the user made among
the computed paths. (It is reasonable to assume that routes picked by the user
are preferred to others, and these observations could contribute to the learning
problem.)
Also interesting is to introduce traffic information into the planner
to support real-time multi-agent concurrent trip planning.

\bibliographystyle{aaai}
\bibliography{refs}

\end{document}

>>>>

Trip planning, as an application of planning and scheduling,
has seen substantive implementations by researchers and
developers\cite{bast2015route}.
Some of the planning systems are \tit{multi-modal}; that is,
combining distinct transportation modes, the trip planners
compute optimal routes from sources to destinations.
This notion of ``optimality" generally refers to the computed routes
having minimal total time or total fare.

@@
